---
title: "3. Sensitivity analysis: the causal model and delta adjustment"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{3. Sensitivity analysis: the causal model and delta adjustment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE,
  fig.width = 7, fig.height = 4.5
)
library(RefBasedMI)
has_mice <- requireNamespace("mice", quietly = TRUE)

quiet <- function(expr) {
  out <- NULL
  utils::capture.output(suppressMessages(out <- expr))
  out
}
```

Sensitivity analyses ask how conclusions change as the post-deviation assumption
is varied. `RefBasedMI` supports two complementary approaches: the **causal
model**, which interpolates between jump-to-reference and copy-increments-in-
reference, and **delta adjustment**, which applies a direct offset for
tipping-point analysis. We use the `asthma` trial throughout, with arm 1 as the
reference and arm 2 as the active arm.

## The causal model

The causal model of White, Royes and Best (2020) retains, at lag `k` after
discontinuation, a fraction `K0 * K1^k` of the last on-treatment effect. The two
constants have a clear interpretation:

- `K0` is the fraction of the treatment effect retained immediately after
  discontinuation;
- `K1` (between 0 and 1) is the per-period decay of that retained effect.

The limits recover the standard methods: `K0 = 1, K1 = 0` is **J2R** (the effect
vanishes immediately) and `K0 = 1, K1 = 1` is **CIR** (the effect is fully
retained).

```{r causal-grid}
causal_run <- function(K1) {
  cl <- bquote(RefBasedMI(
    data = asthma, depvar = fev, treatvar = treat, idvar = id, timevar = time,
    covar = base, method = "Causal", reference = 1, K0 = 1, K1 = .(K1),
    M = 10, seed = 2024
  ))
  quiet(eval(cl))
}

K1_grid <- c(0, 0.25, 0.5, 0.75, 1)
wk12 <- sapply(K1_grid, function(k) {
  out <- causal_run(k)
  d <- out[out$.imp > 0 & out$time == 12 & out$treat == 2, ]
  mean(d$fev)
})
names(wk12) <- paste0("K1=", K1_grid)
round(wk12, 3)
```

```{r causal-plot}
plot(K1_grid, wk12, type = "b", pch = 19, xlab = expression(K[1]),
     ylab = "mean imputed week-12 FEV (active arm)",
     main = "Causal model: J2R (K1=0) to CIR (K1=1)")
```

As `K1` increases from 0 to 1, the imputed active-arm outcome rises from the J2R
value toward the CIR value, tracing the continuum of maintained effect.

## Delta adjustment and tipping-point analysis

Delta adjustment adds a fixed offset to every post-discontinuation imputed value.
By making the offset progressively more unfavourable to the active arm, we can
find the point at which a significant treatment effect is overturned -- a
*tipping-point* analysis.

```{r delta-tipping, eval = has_mice}
treatment_effect <- function(delta_value) {
  out <- quiet(RefBasedMI(
    data = asthma, depvar = fev, treatvar = treat, idvar = id, timevar = time,
    covar = base, method = "J2R", reference = 2,
    delta = c(delta_value, delta_value, delta_value, delta_value),
    dlag = c(1, 1, 1, 1), M = 10, seed = 2024
  ))
  fit <- with(mice::as.mids(out),
              lm(fev ~ factor(treat) + base, subset = (time == 12)))
  est <- summary(mice::pool(fit))
  est[est$term == "factor(treat)2", c("estimate", "p.value")]
}

deltas <- c(0, -0.25, -0.5, -0.75, -1)
tip <- do.call(rbind, lapply(deltas, treatment_effect))
tip <- cbind(delta = deltas, tip)
round(tip, 4)
```

```{r delta-note, eval = !has_mice, echo = FALSE, results = "asis"}
cat("> The tipping-point table requires the **mice** package, which is not",
    "installed, so this chunk is not evaluated.\n")
```

Reading down the table, the estimated active-versus-reference difference at week
12 shrinks (and its p-value grows) as the delta penalty increases, showing how
robust the primary conclusion is to unfavourable departures from J2R.

## Interim (non-monotone) missingness

A missing value that is followed by a later observed value is *interim* rather
than post-discontinuation. `RefBasedMI` always imputes interim values under MAR,
regardless of the chosen reference-based `method`; only monotone, post-
discontinuation missingness uses the reference-based assumption. This matches the
behaviour of the Stata *mimix* program and needs no special handling.

## References

- White I R, Royes J, Best N (2020). A causal modelling framework for
  reference-based imputation and tipping point analysis. *Journal of
  Biopharmaceutical Statistics* 30(2), 334--350.
