---
title: "2. The reference-based imputation methods"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{2. The reference-based imputation methods}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE,
  fig.width = 7, fig.height = 4.5
)
library(RefBasedMI)

# splice a literal method into the call (RefBasedMI captures `method` by NSE)
impute_method <- function(m) {
  cl <- bquote(RefBasedMI(
    data = asthma, depvar = fev, treatvar = treat, idvar = id, timevar = time,
    covar = base, method = .(m), reference = 1, M = 10, seed = 2024
  ))
  out <- NULL
  utils::capture.output(suppressMessages(out <- eval(cl)))
  out
}

# mean outcome trajectory by arm and time, averaged over the imputed datasets
arm_means <- function(out) {
  d <- out[out$.imp > 0, ]
  stats::aggregate(fev ~ treat + time, data = d, FUN = mean)
}
```

This vignette explains what each method assumes about patients after they
deviate from randomised treatment, and shows the consequences on the `asthma`
trial, where the outcome `fev` is measured at weeks 2, 4, 8 and 12 and arm 1 is
the reference (control).

## The assumptions

For a patient who discontinues treatment, the methods differ only in the mean of
the conditional distribution from which post-deviation values are drawn:

| Method | Post-deviation mean | Reference arm |
|--------|---------------------|---------------|
| **MAR**    | continues on the patient's own arm | not required |
| **J2R**    | jumps to the reference arm mean at the first missing visit | required |
| **CR**     | equals the reference arm mean at every visit | required |
| **CIR**    | keeps the increment gained on treatment, then accrues reference increments | required |
| **LMCF**   | holds the last observed mean constant | not required |

MAR is the least conservative (the treatment benefit is assumed to persist);
CR and J2R are the most conservative for an effective experimental arm.

## Observed mean trajectories

```{r observed}
obs <- stats::aggregate(fev ~ treat + time, data = asthma, FUN = mean)
plot(NULL, xlim = c(2, 12), ylim = range(obs$fev), xlab = "week",
     ylab = "mean FEV (observed)", main = "Observed means by arm")
for (a in sort(unique(obs$treat))) {
  d <- obs[obs$treat == a, ]
  lines(d$time, d$fev, type = "b", lty = a, pch = 19, col = a)
}
legend("topright", legend = paste("arm", sort(unique(obs$treat))),
       lty = sort(unique(obs$treat)), col = sort(unique(obs$treat)), pch = 19)
```

## Imputed trajectories under each method

We impute the active arm (arm 2) under each method, with arm 1 as reference, and
plot the imputed mean trajectory (averaged over 20 imputations).

```{r trajectories}
methods <- c("MAR", "J2R", "CR", "CIR", "LMCF")
fits <- lapply(methods, function(m) arm_means(impute_method(m)))
names(fits) <- methods

active <- lapply(fits, function(f) f[f$treat == 2, ])
yr <- range(unlist(lapply(active, function(d) d$fev)))
plot(NULL, xlim = c(2, 12), ylim = yr, xlab = "week",
     ylab = "mean imputed FEV, active arm",
     main = "Active-arm mean by method (reference = arm 1)")
for (i in seq_along(active)) {
  d <- active[[i]]
  lines(d$time, d$fev, type = "b", pch = 19, col = i, lty = i)
}
legend("bottomleft", legend = methods, col = seq_along(methods),
       lty = seq_along(methods), pch = 19, bty = "n")
```

Because most missingness in `asthma` is late dropout, the methods agree at early
visits and fan out by week 12: MAR keeps the active arm highest, while CR and J2R
pull it toward the (lower) reference arm.

## Numerical summary at the final visit

```{r summary-table}
wk12 <- sapply(methods, function(m) {
  f <- fits[[m]]
  f$fev[f$treat == 2 & f$time == 12]
})
round(wk12, 3)
```

## Choosing a method

- Use **MAR** as the primary analysis only if a persisting treatment effect after
  discontinuation is defensible.
- **J2R** and **CR** encode an immediate loss of benefit and are common
  conservative primary or sensitivity analyses for superiority trials.
- **CIR** is intermediate: the benefit already accrued is retained but no further
  benefit is gained.
- **LMCF** is rarely a primary analysis but is useful for comparison.
- The **causal model** provides a continuum between J2R and CIR; see
  `vignette("causal-and-delta")`.
