---
title: "1. Getting started with RefBasedMI"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{1. Getting started with RefBasedMI}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
library(RefBasedMI)
# the imputation routines print progress; silence it for the vignette
impute <- function(...) {
  out <- NULL
  utils::capture.output(suppressMessages(out <- RefBasedMI(...)))
  out
}
```

## Background

In a longitudinal clinical trial, patients who deviate from the protocol (for
example by discontinuing randomised treatment) often have missing outcomes
afterwards. A primary analysis under "missing at random" (MAR) assumes those
patients would have continued along their own arm's trajectory. **Reference-based
multiple imputation** instead encodes assumptions in which deviating patients
resemble a *reference* arm — typically control — which is usually more plausible
and more conservative for an active treatment.

`RefBasedMI` implements these methods, following Carpenter, Roger and Kenward
(2013) and the causal model of White, Royes and Best (2020). It is an R port of
the Stata program *mimix*.

The reference-based methods differ in how the post-deviation mean is constructed:

| Method | Assumption after deviation | Reference arm |
|--------|----------------------------|---------------|
| MAR    | continues on own arm        | not required  |
| J2R    | jumps immediately to the reference mean | required |
| CR     | follows the reference arm throughout | required |
| CIR    | keeps the gained increment, then accrues reference increments | required |
| LMCF   | last observed mean carried forward | not required |
| Causal | maintains a decaying fraction of the treatment effect | required |

## Quick start

The bundled `asthma` trial has the outcome `fev` measured at weeks 2, 4, 8 and
12, with a baseline covariate `base`. We impute under jump-to-reference with
control (arm 1) as reference, creating `M = 5` completed datasets.

```{r quick-start}
imp <- impute(
  data = asthma, depvar = fev, treatvar = treat, idvar = id, timevar = time,
  covar = base, method = "J2R", reference = 1, M = 5, seed = 101
)
dim(imp)
head(imp[, c(".imp", "id", "treat", "time", "fev")])
```

The result is in long format: the original data (`.imp = 0`) stacked above the
`M` imputed datasets (`.imp = 1, ..., M`). Observed values are unchanged; only
the post-deviation missing values are filled in. `summary()` reports the run
settings and confirms that every imputed dataset is complete:

```{r summary}
summary(imp)
```

### Analysis with Rubin's rules

The output is designed to be turned into a `mids` object and pooled. The
exported `as_mids()` helper wraps `mice::as.mids()`:

```{r analysis, eval = requireNamespace("mice", quietly = TRUE)}
fit <- with(as_mids(imp), lm(fev ~ factor(treat), subset = (time == 12)))
summary(mice::pool(fit))
```

## Comparing methods

Because the methods differ only in their assumption about deviating patients, it
is instructive to compare the imputed week-12 means across methods using the same
seed.

```{r compare}
# RefBasedMI captures `method` by non-standard evaluation, so splice the literal
# method string into the call with bquote() rather than passing a loop variable.
impute_method <- function(m) {
  cl <- bquote(RefBasedMI(
    data = asthma, depvar = fev, treatvar = treat, idvar = id, timevar = time,
    covar = base, method = .(m), reference = 1, M = 5, seed = 101
  ))
  out <- NULL
  utils::capture.output(suppressMessages(out <- eval(cl)))
  out
}

methods <- c("MAR", "J2R", "CIR", "CR")
wk12 <- sapply(methods, function(m) {
  out <- impute_method(m)
  mean(out$fev[out$.imp > 0 & out$time == 12 & out$treat == 2])
})
round(wk12, 3)
```

The reference-based methods (J2R, CIR, CR) generally pull the active arm's
imputed outcomes toward the control arm relative to MAR.

## The causal model

The causal model interpolates between J2R and CIR. The maintained treatment
effect after deviation is a fraction `K0` of the last on-treatment effect,
decaying geometrically by `K1` each period: `K0 = 1, K1 = 0` reproduces J2R and
`K0 = 1, K1 = 1` reproduces CIR.

```{r causal}
acu <- impute(
  data = acupuncture, depvar = head, treatvar = treat, idvar = id, timevar = time,
  covar = head_base, method = "Causal", reference = 1, K0 = 1, K1 = 0.5,
  M = 5, seed = 101
)
dim(acu)
```

## Delta adjustment

Delta adjustment adds a user-specified increment to post-deviation imputations
(but not to interim missing values), for tipping-point and sensitivity analyses.
`delta` gives the per-period increments and `dlag` weights them. A constant shift
of `-1` at the first post-deviation visit, accruing thereafter, is:

```{r delta}
impD <- impute(
  data = asthma, depvar = fev, treatvar = treat, idvar = id, timevar = time,
  covar = base, method = "J2R", reference = 2,
  delta = c(-1, 0, 0, 0), dlag = c(1, 1, 1, 1), M = 5, seed = 101
)
dim(impD)
```

## Interim (non-monotone) missingness

A missing value that is followed by a later observed value is treated as
*interim* and imputed under MAR regardless of the chosen reference-based method;
only post-discontinuation (monotone) missingness uses the reference-based
assumption. This matches the *mimix* behaviour and requires no special handling
from the user.

## Choosing M, burnin, and the seed

The examples in this vignette use `M = 5` (and `M = 2` in the help pages) so they
run quickly, but real analyses need more.

- **Number of imputations (`M`).** The Monte Carlo error of a pooled estimate
  falls as `M` grows. A common rule of thumb is to take `M` at least as large as
  the percentage fraction of missing information, which in trials with
  substantial post-deviation missingness often means 50 or more. Too small an `M`
  leaves the standard errors themselves noisy.
- **Burn-in (`burnin`) and spacing (`bbetween`).** Each arm's multivariate normal
  parameters are drawn by the *norm2* MCMC sampler. `burnin` is the number of
  iterations discarded before the first draw; `bbetween` (defaulting to `burnin`)
  is the number of iterations between successive imputations, so that the `M`
  parameter draws are approximately independent. The defaults are generous for
  the small number of repeated measures typical of these trials; increase them if
  a trial has many time points or the sampler reports a near-boundary solution.
- **Reproducibility (`seed`).** `seed` defaults to `NULL`, so imputations are
  random each run. Pass an integer for a reproducible result.

A simple check that `M` is large enough is to repeat the whole imputation under a
few different seeds and confirm that the pooled estimate barely moves:

```{r seed-check, eval = requireNamespace("mice", quietly = TRUE)}
wk12_estimate <- function(seed) {
  imp <- impute(
    data = asthma, depvar = fev, treatvar = treat, idvar = id, timevar = time,
    covar = base, method = "J2R", reference = 1, M = 20, seed = seed,
    verbose = FALSE
  )
  fit <- with(as_mids(imp), lm(fev ~ factor(treat), subset = (time == 12)))
  unname(mice::pool(fit)$pooled[["estimate"]][2])
}
round(vapply(c(101, 202, 303), wk12_estimate, numeric(1)), 3)
```

If those estimates differ materially, increase `M`.

## Other practical points

- **Baseline.** Supply the baseline outcome as a covariate (`covar`) rather than
  as an outcome value, so that no treatment effect is assumed at baseline. Factor
  covariates are accepted and expanded internally.
- **Silencing progress.** Pass `verbose = FALSE` to suppress the per-pattern
  progress messages, which is convenient inside loops such as the seed check
  above.
- **Individual-specific methods.** Assigning a different method per patient via
  `methodvar`/`referencevar` is not supported in this release. To vary the method
  across subgroups, impute each subgroup separately and combine the results.

## References

- Carpenter J R, Roger J H, Kenward M G (2013). Analysis of longitudinal trials
  with protocol deviation. *Journal of Biopharmaceutical Statistics* 23(6),
  1352--1371.
- White I R, Royes J, Best N (2020). A causal modelling framework for
  reference-based imputation and tipping point analysis. *Journal of
  Biopharmaceutical Statistics* 30(2), 334--350.
